Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation

### 2021-03-15 manuscript.Rmd
content

A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue. Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation. In practice, two alternative strategies for capnography alarm management are typically used. The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (15 seconds). The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (30 seconds).

It is possible that capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged. Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.

@@

The prediction goal was to classify apneic events at the 15-second time-point as being either short (i.e. terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer). The prediction algorithm was compared against a default alarm setting for capnography monitors, which triggers an alarm at 30 seconds of apnea.

@@

@@

@@

#### Modelling

>>>>

A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue. Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation. In practice, two alternative strategies for capnography alarm management are typically used. The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (e.g. 15 seconds). The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (e.g. 30 seconds).

It is possible that capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged. Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]

@@

The prediction goal was to classify apneic events at the 15-second time-point as being either short (i.e. terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer). The prediction algorithm was compared against typical default alarm settings for capnography monitors. These are the aggressive strategy, which triggers an alarm after 15 seconds of apnea, and the conservative strategy, which triggers an alarm after 30 seconds of apnea.

@@

Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, dose and type of sedation and analgesia administered) variables were used as predictors. Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation. Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events. A total of 18 variables were used.

@@

Analyses were performed using R[@rcore2021]. Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>. Split into a training and test set containing 75% and 25% of the data, respectively.

@@

#### Modelling
Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.
- Normalized predictors
- interaction terms
